\documentclass[12pt]{ucsddissertation}
% mathptmx is a Times Roman look-alike (don't use the times package)
% It isn't clear if Times is required. The OGS manual lists several
% "standard fonts" but never says they need to be used.
\usepackage{mathptmx}
\usepackage[NoDate]{currvita}
\usepackage{array}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}
\usepackage{microtype}
\usepackage[breaklinks=true,pdfborder={0 0 0}]{hyperref}
\usepackage{graphicx}
\AtBeginDocument{%
	\settowidth\cvlabelwidth{\cvlabelfont 0000--0000}%
}

% OGS recommends increasing the margins slightly.
\increasemargins{.1in}

% These are just for testing/examples, delete them
\usepackage{trace}
%\usepackage{showframe} % This package was just to see page margins
\usepackage[english]{babel}
\overfullrule5pt
% ---

% Packages specific to this dissertation
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{contour}
\usepackage{amsthm}
\usepackage{xcolor}
\usepackage{bussproofs}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{macros}
\usepackage{xspace}
\usepackage{adjustbox}
%\usepackage{gensymb}
%\usepackage{pgfplots}
\usepackage{etex}
\usepackage{todonotes}
\usepackage{mfirstuc}
\usepackage{changebar}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{thmtools}
\usepackage{thm-restate}
\usepackage{bibentry}
\nobibliography*

\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usetikzlibrary{patterns}
\usetikzlibrary{decorations}
\usetikzlibrary{arrows}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{positioning}

\declaretheorem[name=Theorem,numberwithin=chapter]{theorem}
\newtheorem{lemma}{Lemma}[chapter]
\declaretheorem[name=Definition,numberwithin=chapter]{definition}
\newtheorem{assumption}{Assumption}[chapter]

% Required information
\title{Verification of Sampled-Data Systems using Coq}
\author{Daniel Ricketts}
\degree{Computer Science}{Doctor of Philosophy}
% Each member of the committee should be listed as Professor Foo Bar.
% If Professor is not the correct title for one, then titles should be
% omitted entirely.
\chair{Professor Sorin Lerner}
% Your committee members (other than the chairs) must be in alphabetical order
\committee{Professor Samuel Buss}
\committee{Professor William Griswold}
\committee{Professor Ranjit Jhala}
\committee{Professor Todd Millstein}
\degreeyear{2017}

% Start the document
\begin{document}
% Begin with frontmatter and so forth
\frontmatter
\maketitle
\makecopyright
\makesignature
% Optional
\begin{dedication}
\setsinglespacing
\raggedright % It would be better to use \RaggedRight from ragged2e
\parindent0pt\parskip\baselineskip
\centering
\emph{To my wife, parents, and brother, for always believing in me}
\end{dedication}

% Next comes the table of contents, list of figures, list of tables,
% etc. If you have code listings, you can use \listoflistings (or
% \lstlistoflistings) to have it be produced here as well. Same with
% \listofalgorithms.
\tableofcontents
\listoffigures
\listoftables

% Your fancy acks here. Keep in mind you need to ack each paper you
% use. See the examples here. In addition, each chapter ack needs to
% be repeated at the end of the relevant chapter.
\begin{acknowledgements}
I owe a tremendous debt to my adviser Sorin Lerner for supporting me both
technically and otherwise throughout a long PhD. From taking me on as an
untested student switching research areas to letting me spend half of grad
school in Germany with my wife, he has always put my interests first.

A special thanks also goes to my original adviser Mohan Paturi, who always
gave me the freedom to pursue what I enjoyed and supported me when
switching to a new research area and adviser.

It was a great experience working with Gregory Malecha, who showed me the
importance of technical simplicity and abstraction and who convinced me
that everything is an adjoint functor. Don Jang's always positive mood and
ability to finish any task with an impending deadline are an inspiration.
Zachary ``Danger'' Tatlock's excitement for research, outgoing nature, and
ability to bounce back from any setback are a model to emulate, as are his
poncho-based outfits. Valentin Robert helped introduce me to the world of
verification and always kept a relaxed attitude, no matter the situation.

I very much enjoyed working with the Ardupilot developers, particularly
Leonard Hall, who always had time for a fun and enlightening conversation.

One of my earliest collaborators, Moshe Hoffman, showed me what passion for
research looks like, while Petros Mol was always there to re mind me that
there is more to life than work.

I want to thank Wilson Lian for going through le tour de grad school with
me, for all the deep thought-filled conversations, the trap chair-based
entertainment, and for always being a supportive and understanding friend.

I also want to thank Matt Der for being an honest and enthusiastic friend,
and for always prioritizing soccer training over research so that we may
guide the US to a World Cup victory.

I could not have done this without my parents, who taught me to persevere
through all challenges, and my brother Scott, who always has my back and
showed me that you should never worry about what others think.

And most importantly, I want to thank my wife Maja for being a never ending
source of love and support through the highs and lows of Coq, for never
letting me get too frustrated, and for always being proud of me no matter
what.

Chapter~\ref{chap:memo15}, \memoack{}

Chapter~\ref{chap:emsoft16}, \emsoftack{}

Chapter~\ref{chap:exp-smpl}, \expsmplack{}

Chapter~\ref{chap:related}, includes and expands upon material selected
from the above works.
\end{acknowledgements}

% Stupid vita goes next
\begin{vita}
\noindent
\begin{cv}{}
\begin{cvlist}{}
\item[2009] Bachelor of Science, Brown University
\item[2009-2010] Software Engineer, Inria
\item[2011] Intern, Telef\'onica Research
\item[2012] Intern, Facebook
\item[2010--2017] Research Assistant, University of California, San
Diego
\item[2017] Doctor of Philosophy, University of California, San Diego
\end{cvlist}
\end{cv}

% This puts in the PUBLICATIONS header. Note that it appears inside
% the vita environment. It is optional.
\publications
\noindent\bibentry{ricketts2016composition}
\\ \\
\noindent\bibentry{ricketts2015memocode}
\\ \\
\noindent\bibentry{ricketts14reflex}
\\ \\
\noindent\bibentry{cousineau2012tla+}
\\ \\
\noindent\bibentry{dalkiran2012common}

\end{vita}

% Put your maximum 350 word abstract here.
\begin{dissertationabstract}
Due to their safety-critical nature, cyber-physical systems (CPS) demand
the most rigorous verification techniques. However, the complexity of the
domain puts many cyber-physical systems outside the scope of automated
verification tools. Formal deductive proofs hold the potential to verify
virtually any property of any system, but proofs for practical
cyber-physical systems often require an impractical amount of manual
effort. This proof burden can be mitigated by capturing common reasoning
patterns in powerful higher-order proof rules. Existing work has focused on
proof rules applicable to arbitrary hybrid systems (a formal model for
CPS), but many systems actually fall into more constrained classes. One
such class of systems are called sampled-data systems, in which a discrete
controller runs periodically. In this dissertation, we complement general
hybrid system proof rules with a series of rules that leverage the
particular structure of sampled-data systems. We demonstrate the
applicability of these rules on the double integrator, an important model
in robotic and vehicle systems. All work is formalized in the Coq proof
assistant, whose expressive logic is crucial to maintaining soundness while
applying domain-specific proof rules for sampled-data systems. Finally, we
experimentally evaluate our results by implementing verified controllers on
a quadcopter and conducting flight tests.
\end{dissertationabstract}

% This is where the main body of your dissertation goes!
\mainmatter

% Optional Introduction
\begin{dissertationintroduction}
Errors in cyber-physical systems (CPS) can lead to disastrous consequences,
including loss of life.  These consequences mean that such systems demand
the most rigorous verification techniques.  There has been a variety of
work on developing fully-automated tools for verification of cyber-physical
systems~\cite{PHAVerSTTT08,HyTechCAV97}, but due to the complexity of the
domain, these tools are only able to verify particular classes of systems
and properties.  On the other hand, all cyber-physical systems are in range
for deductive verification in a proof assistant, at least in theory.  In
this technique, a user writes a formal model of a system in the language of
the proof assistant and then interactively proves it correct. However, one
of the typically-stated drawbacks of verification in proof assistants is
the extremely high manual labor cost required to produce these proofs.

Mitigating this manual proof burden requires powerful higher-order proof
rules that capture common proof strategies. Prior work in deductive
verification has approached this task by designing general-purpose,
complete proof calculi for hybrid
systems~\cite{Platzer15substitution,LiuHHL10,WangHHL2015}. Hybrid systems
are CPS models comprised of both a discrete component (e.g. control
software) and a continuous component (the physical world). While powerful,
the generality of the proof calculi prevents them from leveraging
particular common characteristics of cyber-physical systems. For example,
no proof rule in~\cite{Platzer15substitution} assumes that the time between
executions of a controller is bounded because this is not true of all
hybrid systems. However, this is true of many realistic hybrid systems, and
proofs about such systems will tend to follow a similar proof
structure. Thus, it is beneficial to complement general hybrid system proof
rules with domain specific proof rules that capture common reasoning
patterns.

This dissertation presents and applies a series of proof rules that capture
common reasoning patterns in the important domain of \emph{periodic
  sampled-data systems}~\cite{chen1995sampled}. In such a system, there is
a digital controller that runs periodically. In between executions of the
controller, the system evolves according to continuous physical
dynamics. Many modern cyber-physical systems fit into this domain. The
remainder of the introduction provides an overview of the proof rules. In
general, the rules seek to leverage timing characteristics of systems and
improve modularity of reasoning.

In \textbf{Chapter~\ref{chap:memo15}} we present two rules: one for
verifying a single sampled-data component under assumption on the
environment and another for composing such a component with another that
satisfies this assumption. The first rule decomposes verification into a
property of the discrete controller and the continuous dynamics,
automatically handling the fact that the time between executions of the
controller is bounded. This tailored decomposition eliminates some of the
basic tedious manipulation common to every sampled-data system, allowing
one to focus on the application specific aspects of verification. The
second proof rule allows for component composition with non-cyclical
assumptions -- that is, a component $C1$ can guarantee an invariant $Q$
while assuming and invariant $P$ guaranteed by $C2$. However, $C2$ cannot
rely on the invariance of $P$ when guaranteeing $Q$. In spite of this
restriction, we show that such a rule has important applications for
verifying controllers in the presence of sensor error and delay.

Next, \textbf{Chapter~\ref{chap:emsoft16}} presents a general framework for
building sampled-data systems \emph{modularly}. This framework differs from
the above composition approach by requiring that each component provide a
stronger interface. In particular, rather than proving invariance of a
property, each component provides preservation of an inductive invariant,
and a notion of progress of the system under that inductive invariant. This
stronger interface comes at a minor cost while proving two important
benefits. First, it allows for cyclic dependencies between sampled-data
components, thus removing the restriction from
Chapter~\ref{chap:memo15}. Second, it allows us to explore a richer set of
operators for modularly building and verifying sampled-data components,
namely substitution, conjunction, and disjunction. It is this second
benefit that we explore thoroughly by applying our framework to build
verified 3-dimensional geofences for a UAV. We show that our theory can
handle the important situation in which different components output to the
same set of actuators, as exemplified by the geofence application.

Finally, in \textbf{Chapter~\ref{chap:exp-smpl}}, we revisit verification
of a single sampled-data system component. Contrary to our prior
approaches, we began by building a geofence that was good enough to be
adopted by the popular open source UAV autopilot called
Ardupilot~\cite{ardupilot}. After building such a module (now available in the
latest Ardupilot release), we attempted to formally verify a component of
it in Coq, particularly the logic that prevents the vehicle from violating
a boundary in a single spatial dimension. Similar logic is present in the
atomic components from Chapters~\ref{chap:memo15} and~\ref{chap:emsoft16},
but realistic performance requirements for Ardupilot resulted in
considerably more complicated control logic. This additional complexity
demanded better proof rules, and we built rules that improve upon the state
of the art in formal verification in two ways.

First, deductive techniques typically involve some continuous analogue of
induction, e.g. differential induction~\cite{Platzer10DAL} or barrier
certificates~\cite{prajna04barrier}. Recent work from the control theory
community~\cite{kong2013barrier,xu15barrier,nguyen16barrier} has produced a
new version of barrier certificates that are less conservative for closed
properties. We provide the first implementation of this approach in a
formal verification context, and demonstrate its ease of use on a component
of the Ardupilot geofencing module.

Second, control systems are often designed under the assumption that
controllers run continuously, while the actual implementation is typically
a sampled-data system. System designers can compensate for this (and other)
approximations by adding a safety ``buffer'' to the system. For example,
the Ardupilot geofence module stops the vehicle 1 meter prior to the actual
safety boundary. We developed a proof rule that formalizes this design
approach by decomposing verification into a condition on the continuous
time approximation and another on the approximation error.  This rule
allows one to perform the majority of reasoning in a purely continuous
model using powerful techniques resulting from over a century of control
theory research~\cite{kong2013barrier,xu15barrier,nguyen16barrier}.

As already mentioned, our running application in this work is a geofencing
controller for UAVs, an important application due to their potential safety
threat combined with widespread use by hobbyists and businesses alike. We
formally model such systems using the double integrator from control
theory, in which the controller affects the acceleration (second
derivative) of position. This is an approximation of reality, as the actual
quadcopter controller affects the angular torque of the vehicle. The
angular torque is related to the second derivative of position via complex
attitude dynamics. Although our model is an approximation, the double
integrator is a benchmark problem for robotics, vehicles, and other systems
that involve a point mass moving in space~\cite{rao2001naive}. Thus, the
double integrator serves as a canonical example for evaluating the
expressiveness of the proof rules presented in this dissertation.

Since we use an approximate model of UAV dynamics, we would like to ensure
that the controllers we build and verify are not toys. We have attempted to
justify the realistic nature of our work by flying the controllers we
verify on an actual quadcopter. Throughout this dissertation, we discuss
lessons learned from this experience.

Rigorous verification requires that results be mechanically checked in some
way. Rather than implementing a standalone tool for this task, we performed
all verification within the foundational Coq proof assistant. Previous
work~\cite{yang2011understanding-compiler-bugs} has empirically
demonstrated that foundationally verified systems are highly
reliable. While important, we would like to emphasize a less discussed
benefit of higher-order proof assistants like Coq. A standalone domain
specific verification tool would require adding each proof rule as an
axiom. Moreover, a new proof rule might have side conditions not
expressible in the tool's logic, requiring custom reasoning to handle these
side conditions. In this context, the axiom and associated custom reasoning
have the potential to compromise soundness.

On the other hand, the \emph{expressiveness} of Coq allows one add new
proof rules as theorems that are formally proven within Coq's logic. This
means that one can extend any given set of general proof rules in Coq
(e.g. general proof calculi for hybrid systems) with powerful domain
specific proof rules (e.g. our sampled-data system rules), without
compromising soundness. Such an extension improves verification
productivity by ensuring that a user can apply the right domain-specific
proof rule for his or her application while still being able to depend on
the above mentioned reliability guarantees.
\thesis
It is thus the view of the dissertation author that the expressive
framework of a higher-order proof assistant is crucial to scaling formal
verification to realistic cyber-physical systems, and we hope that the
proof rules and associated applications in this document provide evidence
for this claim.

\end{dissertationintroduction}

\chapter{Preliminaries}
\label{chap:prelim}
\input{prelim}

\chapter{Discrete Induction and Basic Composition}
\label{chap:memo15}
\input{memo15}

\chapter{Modular Verification}
\label{chap:emsoft16}
\input{emsoft16}

\chapter{Barrier Certificates}
\label{chap:exp-smpl}
\input{exp-smpl}

\chapter{Related Work}
\label{chap:related}
\input{related}

\chapter{Conclusions and Future Work}
\label{chap:conclusion}
\input{conclusion}

%\appendix

% Stuff at the end of the dissertation goes in the back matter
\backmatter
\bibliographystyle{plain} % Or whatever style you want like plainnat
\bibliography{bib,platzer,intro}

\end{document}
