\documentclass[10pt]{article}
\usepackage{ucsdcseletterhead}
\usepackage{parskip}
\usepackage[margin=1in]{geometry}
\usepackage{mathptmx}
\usepackage{courier}
\usepackage[scaled=.92]{helvet}
\usepackage[T1]{fontenc}
\pagestyle{empty}
\begin{document}
\hfill\today

Steven Cassedy, Dean\\
Graduate Studies\\
Student Services Center, 4th Floor
UC San Diego\\
La Jolla, CA 92093-0003

\vskip\baselineskip
Dear Dean Cassedy:

We request permission for Daniel Ricketts to use the following publications
in his doctoral dissertation. Mr.\ Ricketts was the principle
researcher/author on this paper.

Ricketts, Daniel; Malecha, Gregory; Lerner, Sorin. ``Modular deductive
verification of sampled-data systems'' Proceedings of the 13th
International Conference on Embedded Software, pages 17:1-17:10.

Ricketts, Daniel; Malecha, Gregory; Alvarez, Mario M.; Gowda, Vignesh;
Lerner, Sorin. ``Towards Verification of Hybrid Systems in a Foundational
Proof Assistant'', 2015 ACM/IEEE International Conference on Formal Methods
and Models for Codesign (MEMOCODE), pages 248-257, Sept 2015.

We request permission for Daniel Ricketts to use material currently being
prepared for submission for publication. Mr. Ricketts was the principal
researcher/author on this paper.

Ricketts, Daniel; Ghaffari, Azad; Abel, Imoleayo; Lerner, Sorin; Krstic,
Miroslav. ``State and Input Constraints for the Double Integrator using
Barrier Certificates.''

\vskip3\baselineskip
\parbox[t]{.5\textwidth}{%
Professor Sorin Lerner\\
Chair of Committee}%
\parbox[t]{.5\textwidth}{%
Daniel Ricketts}
\end{document}
